---
title: "R Notebook"
output: html_notebook
---

# R: Data Structures

## Vectors

Vectors are a list-like data structure that can contain any number of items of the same data type. Vector elements are separated by commas, and the entire vector must be wrapped in `c()`

```{r}
flags <- c(TRUE, FALSE, TRUE, FALSE)


```

### Getting a Vector's Type

You can get the type of the elements inside of a vector using the `typeof()` function

```{r}
nums <- c(1, 2, 3, 4, 5)
typeof(nums)



```

### Getting a Vector's Length

You can get the length of a vector using the `length()` function

```{r}

nums <- c(1, 2, 3, 4, 5)
fruits <- c("apple", "banana", "orange")
flags <- c(TRUE, FALSE, TRUE)


length(nums)    
length(fruits)
length(flags)   

```

### Indexing a Vector

You can get the value at a specific index in your vector by placing the elements position within brackets, ex: `[1]` .

**NOTE: Indexes in R start at 1, NOT 0.**

```{r}
# Note this returns "John", the first element in our Vector

names <- c("John", "Jane", "Doe")

print(names[1])  

```

### Generating a sequence between two numerical values

```{r}
# Generate a sequence from 5 to 15

seq_5_to_15 <- seq(5, 15)
print(seq_5_to_15)


# Generate a sequence with decimals

```

#### Using the seq() function

You can also create sequences with more customization by using the `seq()` function. With this you can set a "step" value (this is the `by` parameter).

```{r}
# Step from 5 to 15 by increments of .3

seq_custom <- seq(from = 5, to = 15, by = 0.3)
print(seq_custom)

```

### Vector Sorting

You can sort vectors using the `sort()` function

```{r}

nums <- c(5, 2, 9, 1, 7)


sorted_nums <- sort(nums)
print(sorted_nums)

```

### Vector Calculations

#### Addition

```{r}
vec1 <- c(1, 2, 3)
vec2 <- c(4, 5, 6)


addition_result <- vec1 + vec2
print(addition_result)

```

#### Subtraction

```{r}

subtraction_result <- vec2 - vec1
print(subtraction_result)

```

#### Multiplication

```{r}

multiplication_result <- vec1 * vec2
print(multiplication_result)

```

#### Division

```{r}

division_result <- vec2 / vec1
print(division_result)

```

### Vector Recycling

If we perform vector calculations on uneven vectors, the elements of the shorter vectors are reused to fill in the remaining length

```{r}

```

## Lists

Lists in R are generic objects that contain an ordered collection of objects. Lists can contain a variety of data types (they are heterogeneous). Lists are created with the `list()` function.

```{r}

my_list <- list(
  name = "John",
  age = 30,
  height = 175.5,
  grades = c(85, 92, 78),
  is_student = TRUE
)

print(my_list)

```

```{r}

my_list <- list(
  name = "John",
  age = 30,
  height = 175.5,
  grades = c(85, 92, 78),
  is_student = TRUE
)

print(my_list)

print(my_list$name)   
print(my_list[[2]])    
print(my_list[[4]])    

```

## Dataframes

Dataframes are generic objects that store tabular data. They are two-dimensional data structures that can store any data-type (they are heterogeneous). Dataframes store lists of vectors of equal length (each vector is, essentially, a row).

Dataframes have the following constraints: 

-   A dataframe must have column names and every row should have a unique name.

-   Each column must have an identical number of items.

-   Each item in a single column must be of the same data type.

-   Different columns may have different data types.

When creating a dataframe, you first create the values for each column, and use these to generate your dataframe. You generate a dataframe using the `data.frame()` function.

```{r}
# Example data for the data frame
student_id <- c(1, 2, 3, 4, 5)
student_name <- c("Alice", "Bob", "Charlie", "David", "Eve")
age <- c(22, 23, 21, 24, 22)
grade <- c("A", "B", "A", "B", "A")

# Create the data frame
student_data <- data.frame(StudentID = student_id,
                           Name = student_name,
                           Age = age,
                           Grade = grade)

# Print the data frame
print("Student Data:")
print(student_data)

```

### str()

You can get the structure of a dataframe using the `str()` function

```{r}
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Using str() on the data frame
str(student_data)

```

### summary()

You can get a summary of the data inside of a dataframe using the `summary()` function. This returns useful information like the type of each column and statistical information like the mean, median, and mode of numerical columns.

```{r}
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Using summary() on the data frame
summary(student_data)

```

### Retrieving data from a dataframe

You can index specific columns from a dataframe using the dataframe variable, then a `$` , then the name of the column.

```{r}
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Accessing specific columns using $
student_names <- student_data$Name
student_grades <- student_data$Grade

# Printing the retrieved columns
print("Student Names:")
print(student_names)

print("Student Grades:")
print(student_grades)

```

You can also index specific rows by calling the name of the dataframe, followed by square brackets and the rows you want to retrieve. The square brackets accept two values, the first is the rows you want, the second is the columns.

```{r}
# Get the first two rows and all of the columns
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Get the first two rows and all columns
first_two_rows <- student_data[1:2, ]

# Print the retrieved data
print("First Two Rows and All Columns:")
print(first_two_rows)

```

```{r}
# Get the first and third rows and all of the columns
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Get the first and third rows and all columns
selected_rows <- student_data[c(1, 3), ]

# Print the retrieved data
print("Selected Rows and All Columns:")
print(selected_rows)

```

```{r}
# Get the first and third rows and the first and third columns

```

### Adding New Columns

A new column can be added to a dataframe by indexing the column using the `$` symbol and setting the value to a vector

```{r}
# Example data frame
student_data <- data.frame(StudentID = c(1, 2, 3, 4, 5),
                           Name = c("Alice", "Bob", "Charlie", "David", "Eve"),
                           Age = c(22, 23, 21, 24, 22),
                           Grade = c("A", "B", "A", "B", "A"))

# Get the first and third rows and the first and third columns
selected_data <- student_data[c(1, 3), c(1, 3)]

# Print the retrieved data
print("Selected Rows and Columns:")
print(selected_data)

```

### Adding New Rows

To add new rows to a dataframe, we need to first create a new dataframe using the same structure as the original, then combine the two using the `rbind()` function

```{r}
# Original data frame
student_data <- data.frame(StudentID = c(1, 2, 3),
                           Name = c("Alice", "Bob", "Charlie"),
                           Age = c(22, 23, 21),
                           Grade = c("A", "B", "A"))

# New rows to add
new_rows <- data.frame(StudentID = c(4, 5),
                       Name = c("David", "Eve"),
                       Age = c(24, 22),
                       Grade = c("B", "A"))

# Add new rows using rbind()
updated_student_data <- rbind(student_data, new_rows)

# Print the updated data frame
print("Updated Student Data:")
print(updated_student_data)

```

## Matrices

A matrix is a two-dimensional assortment of number data points in columns and rows. In matrices columns run horizontally and rows run vertically. To create a matrix you call the `matrix()` function. You'll have to pass in a vector of data to generate the matrix from, the number of rows you want, and the number of columns you want. Remember, matrices are in column-wise order.

The following parameters are available to you using matrices:

-   **data** is the input vector.

-   **nrow** is the number of rows to be created.

-   **ncol** is the number of columns to be created.

-   **byrow** If TRUE then the input vector elements are arranged by row, if FALSE they are arranged by column. This is false by default.

-   **dimnames** is the names assigned to the rows and columns.

```{r}
# Example vector of data
data_vector <- 1:12  # Vector with 12 elements

# Create a matrix with 3 rows and 4 columns, filled column-wise (default)
my_matrix <- matrix(data_vector, nrow = 3, ncol = 4)

# Print the matrix
print("Matrix created:")
print(my_matrix)

```

### Naming Columns and Rows

You can name columns and rows with the optional dimnames parameter

```{r}
# Example vector of data
data_vector <- 1:12  # Vector with 12 elements

# Create a matrix with 3 rows and 4 columns, filled column-wise (default)
my_matrix <- matrix(data_vector, nrow = 3, ncol = 4)

# Print the matrix
print("Matrix created:")
print(my_matrix)

```

### Accessing Matrix Values

Values in a matrix can be accessed by indexing with square brackets.

```{r}
# Get the element in the 2nd row and 4th column
# Example matrix
mat <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
print(mat)

```

You can also get entire rows and columns.

```{r}
# Example matrix
mat <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
print(mat)

```

### Calculations between two matrices

```{r}
# Define matrices A and B
A <- matrix(1:9, nrow = 3, byrow = TRUE)  # 3x3 matrix
B <- matrix(10:18, nrow = 3, byrow = TRUE)  # 3x3 matrix

# Print matrices A and B
print("Matrix A:")
print(A)

print("Matrix B:")
print(B)

```

#### Addition

```{r}
# Define matrices A and B
A <- matrix(1:9, nrow = 3, byrow = TRUE)  # 3x3 matrix
B <- matrix(10:18, nrow = 3, byrow = TRUE)  # 3x3 matrix

# Print matrices A and B
print("Matrix A:")
print(A)

print("Matrix B:")
print(B)
# Addition of matrices A and B
addition_result <- A + B
print("Result of Addition:")
print(addition_result)

```

#### Subtraction

```{r}
# Define matrices A and B
A <- matrix(1:9, nrow = 3, byrow = TRUE)  # 3x3 matrix
B <- matrix(10:18, nrow = 3, byrow = TRUE)  # 3x3 matrix

# Print matrices A and B
print("Matrix A:")
print(A)

print("Matrix B:")
print(B)

# Subtraction of matrix B from matrix A
subtraction_result <- A - B
print("Result of Subtraction:")
print(subtraction_result)

```

#### Multiplication

```{r}
# Define matrices A and B
A <- matrix(1:9, nrow = 3, byrow = TRUE)  # 3x3 matrix
B <- matrix(10:18, nrow = 3, byrow = TRUE)  # 3x3 matrix

# Print matrices A and B
print("Matrix A:")
print(A)

print("Matrix B:")
print(B)

# Element-wise multiplication of matrices A and B
multiplication_result <- A * B
print("Result of Multiplication (Element-wise):")
print(multiplication_result)

```

#### Division

```{r}
# Define matrices A and B
A <- matrix(1:9, nrow = 3, byrow = TRUE)  # 3x3 matrix
B <- matrix(10:18, nrow = 3, byrow = TRUE)  # 3x3 matrix

# Print matrices A and B
print("Matrix A:")
print(A)

print("Matrix B:")
print(B)

# Element-wise division of matrix A by matrix B
division_result <- A / B
print("Result of Division (Element-wise):")
print(division_result)

```

## Arrays

Arrays can store data in more than two dimensions.They are n-dimensional. If we create an array with the dimensions of (5,6,6), it will create 6 matrices with 5 rows and 6 columns. Arrays are created with the `array ()` function. The first argument is the vector of data to display, the second is the desired dimensions for your array.

```{r}
# Example data vector to populate the array
data_vector <- 1:(5 * 6 * 6)  # Create a vector with enough elements for a 5x6x6 array

# Create a 3-dimensional array
my_array <- array(data_vector, dim = c(5, 6, 6))

# Print the array
print(my_array)

```

### Naming Array columns and rows

You can name columns, rows, and matrices using the optional dimnames parameter

```{r}
# Example data for the array
data1 <- matrix(1:9, nrow = 3)
data2 <- matrix(10:18, nrow = 3)
array_data <- array(c(data1, data2), dim = c(3, 3, 2))

# Define names for rows, columns, and matrices
row_names <- c("Row1", "Row2", "Row3")
col_names <- c("ColA", "ColB", "ColC")
mat_names <- c("Matrix1", "Matrix2")

# Create array with named dimensions
named_array <- array(array_data, dim = c(3, 3, 2),
                     dimnames = list(row_names, col_names, mat_names))

# Print the named array
print(named_array)

```

### Accessing Array Elements

You can access array elements with square brackets, then three arguments. The row, the column, then the matrix.

```{r}

arr <- array(c(1:9, 10:18), dim = c(3, 3, 2))

# Get the element in the second row, third column, first matrix
element_1 <- arr[2, 3, 1]
print(element_1)

```

```{r}


second_row_all <- arr[2, , ]
print(second_row_all)

```

```{r}

second_column_all <- arr[, 2, ]
print(second_column_all)


```

```{r}

# Get the second matrix
second_matrix <- arr[,, 2]
print(second_matrix)

```

## Factors

Factors are used to categorize and store data. Factors are primarily useful when working with categorical data. Factors are useful in data analysis for statistical modeling. Factor data is expected to be a vector of limited, unique values.

Factors are created using the `factor()` function and passing in a vector of data.

```{r}
# A factor of eye colors
# Example vector of eye colors
eye_colors <- c("Blue", "Brown", "Green", "Brown", "Blue", "Green", "Blue", "Brown")

# Create a factor from the vector
eye_color_factor <- factor(eye_colors)

# Print the factor and its levels
print(eye_color_factor)
print(levels(eye_color_factor))

```

You can also use the optional levels parameter to specify the order in which the levels should be displayed.

```{r}
# Example vector of eye colors
eye_colors <- c("Blue", "Brown", "Green", "Brown", "Blue", "Green", "Blue", "Brown")

# Define the desired order of levels
desired_levels <- c("Brown", "Blue", "Green")  # Specify the order as desired

# Create a factor with specified levels
eye_color_factor <- factor(eye_colors, levels = desired_levels)

# Print the factor and its levels
print(eye_color_factor)
print(levels(eye_color_factor))

```

## Homework

1.  Create an array of two 3x3 matrices from the given vectors.

```{r}
# Example vectors (you can replace these with your actual vectors)
vec1 <- 1:9
vec2 <- 10:18

# Create matrices from vectors
mat1 <- matrix(vec1, nrow = 3, ncol = 3)
mat2 <- matrix(vec2, nrow = 3, ncol = 3)

# Create an array of matrices
arr <- array(c(mat1, mat2), dim = c(3, 3, 2))

print(arr)

```

2.  In the above matrix that you created, print the value 5 by indexing in to the matrix.

3.  Print the third row of the first matrix.

4.  Print the second column of both matrices.

    ```{r}
    # Accessing the value 5 in the first matrix (indexing starts from 1)
    value_5 <- arr[2, 2, 1]
    print(value_5)

    # Print the third row of the first matrix
    third_row_first_matrix <- arr[3, , 1]
    print(third_row_first_matrix)

    # Print the second column of both matrices
    second_column_both_matrices <- arr[, 2, ]
    print(second_column_both_matrices)


    ```
